Remote sensing riverine gas exchange and estimating carbon efflux using SWOT observations

### 2021-02-18 manuscript_v1.Rmd
manuscript edits

---
title: "Remote sensing riverine gas exchange and carbon efflux using SWOT observations"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
bibliography: References.bib
csl: rse_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##Abstract
Go BIKER!!
400 words
x
x
x
x
x
x
x
x
x

@@

It is infeasible to measure $k_{600}$ directly when working at the drainage-network-scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007]. This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976]. Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by estimates of the *V* of the river channel. Thus, drainage-network-scale GHG emissions estimates are sensitive to their *V* data. This is exasperated in ungauged basins where *V* models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

To circumvent this problem in ungauged basins/at drainage-network-scales, to date researchers have used ‘global-scope’ hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge [@leopoldHydraulicGeometryStream1953a] that are trained on extensive hydraulics datasets. This is the approach used in upscaling *V* to continental/global river networks and has proven useful in those settings [e.g. @horgbyUnexpectedLargeEvasion2019; @borgesVariabilityGasTransfer2004; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a]. However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river [@parkWorldwideVariationsHydraulic1977]. Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters [e.g. @dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a] and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale). A potential alternative to this approach is to directly estimate a river’s hydraulic properties using remote sensing (RS) data.

Remote hydraulic estimation is itself a burgeoning subfield within remote sensing of hydrology and an analogue to the problem presented here is the remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review). There are many ways to perform RSQ, from calibrating hydrologic and hydraulic models [] to methods that use no in situ information ('ungauged approaches') []. Many of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.

SWOT will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle. It will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. In the context of RSQ, a decade of SWOT work has explored the multi-parameter problem of estimating roughness and bathymetry to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms (e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a). They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017]. While the raw performance of McFLIs is lower than the more sophisticated approaches mentioned above, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b]. Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation [@ishitsukaCombiningOpticalRemote2020].

@@

Thus, to advance our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales, we sought to estimate $k_{600}$ using only RS data and no in situ information. We hypothesize that remote sensing $k_{600}$ is possible by coupling 1) hydraulic geometry theory, 2) $k_{600}$ upscaling theory, 3) Bayesian remote sensing (in a manner similar to McFLI), and 4) SWOT data. More specifically, this manuscript aims to answer two questions:
	1) Is ungauged RS of the gas exchange velocity possible?
	2) How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?

To answer these questions, we build a RS algorithm that ingests SWOT data and produces $k_{600}$ estimates (with fully quantified uncertainty within a Bayesian framework) that requires no in situ information about a river’s hydraulic properties. We validate the algorithm on simulated SWOT data for 22 rivers (as SWOT has yet to launch, simulated data is necessary- Section 2.3). We also quantify the influence of the expected measurement errors in SWOT data on our $k_{600}$ estimates. Finally, we compare our algorithm against published methods: we use previously published dissolved carbon dioxide ($CO_2$) data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and published velocity rating curves and in situ discharge records.

##2 Methods
We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’ algorithm. To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that does not rely on a priori knowledge of the river (section 2.1) and then we use that upscaling model to build the actual RS algorithm (section 2.2). Following these methods, the validation setup is described (section 2.3), as is the comparison of estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4). A flowchart detailing the complete workflow is included (Figure 1).

![Figure 1. Flowchart of methodology developed in this chapter. ‘Extended Upscaling Model’ is detailed in section 2.1. ‘BIKER’ is detailed in section 2.2 and ‘Flux Estimation’ is detailed in section 2.4. See section 2.3 for the validation setup.](outputs/flowchart.jpg)

###2.1 Developing an RS-able $k_{600}$ upscaling model
[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient [e.g. @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976]. Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a]’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers. More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low. Their upscaling function is reprinted as equation 2.

@@

While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters. We do not have this a priori knowledge in an ungauged setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width as a proxy for river size and train different upscaling functions for different groups of data with similar river sizes (Figure 2, using data from [@ulsethDistinctAirWater2019a] dataset). In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

@@

This model therefore assigns $a_i$ and $b_i$ a priori using just river width and slope to define *i*. Another implication of this extension of the [@ulsethDistinctAirWater2019a] model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.

@@

Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. We specifically explore three uncertainty scenarios to assess which error source is most influential. Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error. Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes. To do this, we ran three sets of MC simulations on 5,000 sets of field measurements of river channel hydraulics [@brinkerhoffReconcilingAtaStationAtManyStations2019]. Each MC simulation was itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3). Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25. The specifics of the slope-error model used are elaborated in Section 2.3, but in short we account for both the SWOT error tolerances and radar layover error [@durandHowWillRadar2020a]. This resulted in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term is extracted and used for $σ_{upscale}^2$.

With the sampling model (equation 5), prior distributions, and the $σ_{upscale}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is derived. To approximate this distribution, BIKER uses a Markov Chain Monte Carlo (MCMC) algorithm. Because it is written in the Stan probabilistic programming language, BIKER’s MCMC of choice is a Hamiltonian Monte Carlo which reduces computation time relative to other sampling algorithms (https://mc-stan.org/docs/2_26/reference-manual/hamiltonian-monte-carlo.html).

@@

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers. Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no measurement errors) and represent the best-case scenarios of what SWOT will provide to hydrologists. Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations within the SWOT context. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately algins the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4). All told, this yielded 503 sets of SWOT observations to validate BIKER with.

We also introduced measurement error into these simulated SWOT observations to assess the performance degradation due to errors inherent in SWOT measurements. Error in SWOT-measured river heights and water surface slopes will come from tolerances internal to the satellite, i.e. the SWOT error budget, as well as radar layover error. River width errors are assumed here to be neglible [@durandHowWillRadar2020a]. Layover error is the phenomenon when radar returns arrive at the sensor at the same time, generally in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013]. While the SWOT simulator mentioned above is ideal for this analysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT radar layover errors [@durandHowWillRadar2020a]. They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches. The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively. To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from a normal distribution as by equations 7 and 8. Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a]: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$ given our rectangular channel assumptions.

@@

Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself (which can be validated using existing datasets). Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against. This is calculated using equation 9, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area. With this setup, we can directly explore our ability to infer observed *V* and $k_{600}$ from river width and height alone. It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$). Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation. However in practice, it should reflect the total uncertainty calculated in section 2.2.

@@

##3 Results
First, we present the results from the $k_{600}$ upscaling experiment (section 3.1). We then take the best performing upscaling function, implement it within BIKER, and validate it on the 22 SWOT rivers (section 3.2). Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates from the 22 rivers (section 3.3).

###3.1 Validating the $k_{600}$ upscaling model
Figure 3 plots the validations for the five upscaling functions tested. Both the original [@ulsethDistinctAirWater2019a] model, and our RS-able extension, perform the best on this independent set of validation data, with near identical performance. The RS-able [@ulsethDistinctAirWater2019a] model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original [@ulsethDistinctAirWater2019a] model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively). The two power-law based Raymond models ([@raymondScalingGasTransfer2012a] equations 4 and 3) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively). [@raymondScalingGasTransfer2012a] equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset. In fact, it predicts negative $k_{600}$ values and error metrics cannot be calculated because the prediction residuals are not normally distributed. Because we successfully reproduced the [@ulsethDistinctAirWater2019a] model, the rest of this manuscript deals exclusively with this RS-able variant of the model.

>>>>

---
title: "Remote sensing riverine gas exchange and estimating carbon efflux using SWOT observations"
author: "CB Brinkerhoff, CJ Gleason, PA Raymond, MH Harlan"
date: "`r Sys.Date()`"
bibliography: References.bib
csl: rse_citation_style.csl
output:
  word_document: 
      reference_docx: word-styles-reference.docx
---

@@

##Abstract
Go BIKER!!
max 400 words
x
x
x
x
x
x
x
x
x

@@

It is infeasible to measure $k_{600}$ directly when working at the drainage-network-scale and so ‘upscaling functions’ are used that predict $k_{600}$ from readily available river geomorphology [e.g. @oconnorMechanismReaerationNatural1958; @palumboAssessingPerformanceReaeration2014; @raymondScalingGasTransfer2012a]. Many upscaling functions are based on the basic theory that *k* is empirically correlated with the turbulent energy dissipation *e* at the air-water interface [@zappaEnvironmentalTurbulentMixing2007]. This is convenient for upscaling because in rivers and streams *e* can be equated to the turbulent energy dissipation rate *eD* $[L^2/T^3]$ [@ulsethDistinctAirWater2019a; @raymondScalingGasTransfer2012a], which is estimated using easily calculated hydraulic parameters: $eD \approx gSV$ where *g* is gravitational acceleration $[L/T^2]$, *S* is channel slope, and *V* is average flow velocity $[L/T]$ [@tsivoglouTracerMeasurementReaeration1976]. Because *S* is readily available in any hydrographic data product, most efforts to upscale $k_{600}$ to drainage networks are limited by estimates of the *V* of the river channel. Thus, drainage-network-scale GHG emissions estimates are highly sensitive to their *V* data. This is exasperated in ungauged basins where *V* models cannot be trained because no in situ hydraulic information is available and fieldwork is impractical [@gleasonRemoteSensingRiver2020].

To circumvent this problem in ungauged basins/at drainage-network-scales, to date researchers have used ‘global-scope’ hydraulic geometry (HG) models: simple power law relations between hydraulic properties and discharge [@leopoldHydraulicGeometryStream1953a] that are trained on extensive hydraulics datasets. This is the approach used in upscaling *V* to continental/global river networks and has proven useful in those settings [e.g. @borgesVariabilityGasTransfer2004; @horgbyUnexpectedLargeEvasion2019; @lauerwaldSpatialPatternsCO22015; @raymondGlobalCarbonDioxide2013a]. However, it has long been established that HG parameters exhibit high spatial variability across landscapes and vary significantly from river to river [@parkWorldwideVariationsHydraulic1977; @rhodesBfmDiagramGraphical1977]. Because of this, a large body of geomorphology work has attempted to parse out process-based explanations for HG parameters [e.g. @dingmanAnalyticalDerivationAtastation2007b; @parkerPhysicalBasisQuasiuniversal2007b; @singhTHEORIESHYDRAULICGEOMETRY2003a; @fergusonHydraulicsHydraulicGeometry1986a] and an open research question is how to best prescribe a set of HG parameters to a given river (particularly at the global scale). A potential alternative to this approach is to directly estimate a river’s hydraulic properties using remote sensing (RS) data.

Remote hydraulic estimation is itself a burgeoning subfield within remote sensing of hydrology and an analogue to the problem presented here is the remote sensing of river discharge (RSQ- see @gleasonRemoteSensingRiver2020 for a thorough review). There are many ways to perform RSQ, from calibrating to channel hydraulics [e.g. @brakenridgeOrbitalMicrowaveMeasurement2007; @pavelskyUsingWidthbasedRating2014; @pavelskyRemoteSensingSuspended2009; @tarpanelliEstimationRiverDischarge2013] to calibrating hydraulic/hydrologic models [e.g. @bjerklieEstimatingDischargeRivers2005; @chandanpurkarSatelliteReanalysisBasedMass2017; @linGlobalReconstructionNaturalized2019a; @nealDataAssimilationApproach2009] to methods that use no in situ information as inputs to  hydraulic/hydrologic models (termed 'ungauged approaches'). Many of these ungauged approaches were developed in the context of the upcoming NASA/CNES/UKSA/CSA Surface Water and Ocean Topography (SWOT) satellite mission.

SWOT will launch in 2022 and provide the world’s first global measurements of water surface extent and elevation at novel temporal resolutions. SWOT is a wide swath Ka-band radar interferometer and will sample rivers every 1 to 7 days per 21 day repeat cycle. It will measure rivers wider than 100m with a goal of expanding this to rivers at least 50m wide [@biancamariaSWOTMissionIts2016]. In the context of RSQ, a decade of SWOT work has explored the multi-parameter problem of estimating roughness and bathymetry to produce the SWOT discharge product [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @durandEstimatingReachaveragedDischarge2014a; @garamboisInferenceEffectiveRiver2015; @gleasonRetrievalRiverDischarge2014a; @garamboisVariationalEstimationEffective2020; @hagemannBAMBayesianAMHGManning2017a; @larnierRiverDischargeBathymetry2020a; @oubanasDischargeEstimationUngauged2018]. A specific subset of these ungauged RSQ methods are termed ‘McFLIs’ or ‘Mass Conserved Flow Law Inversion’ algorithms [e.g. @andreadisConstrainingAssimilationSWOT2020d; @brinkerhoffConstrainingRemoteRiver2020; @hagemannBAMBayesianAMHGManning2017a; @durandEstimatingReachaveragedDischarge2014a]. They use basic geomorphic theories, rather than full hydrologic and/or hydraulic models, and the concept of ‘prior river knowledge’ to estimate discharge from RS data where not all hydraulic terms are RS-able [@gleasonRemoteSensingRiver2020; @gleasonTrackingRiverFlows2017]. While the raw performance of McFLIs is lower than the more sophisticated approaches mentioned above, they are globally flexible, easily implementable in any river that SWOT can sample, and improve our hydrological understandings of ungauged rivers when little to no information was previously available [@brinkerhoffConstrainingRemoteRiver2020; @durandIntercomparisonRemoteSensing2016b]. Recently, McFLIs have also shown promise in providing additional and beneficial information to traditional hydrologic modeling via data assimilation [@ishitsukaCombiningOpticalRemote2020].

@@

Thus, we hypothesize that SWOT data will improve our understanding of riverine GHG fluxes in ungauged basins and/or at drainage-network-scales, after techniques have been devleoped to ingest its novel measurements of water surface width and height. Therefore, we sought to develop a method to estimate $k_{600}$ using only SWOT measurements and no in situ information. We hypothesize that remote sensing $k_{600}$ is possible by coupling Bayesian remote sensing techniques (in a manner similar to McFLI) and SWOT data with hydraulic geometry and gas exchange theory. More specifically, this manuscript aims to answer two questions:
	1) Is ungauged RS of the gas exchange velocity possible using soon-to-be-available SWOT river measurements?
	2) How might this algorithm influence upscaled estimates of riverine carbon efflux to the atmosphere?

To answer these questions, we develop a RS algorithm that ingests SWOT data and produces $k_{600}$ estimates and associated uncertainties that requires only water surface width and slope and therefore no in situ information about a river’s hydraulic properties. We validate the algorithm on simulated SWOT data for 22 rivers (as SWOT has yet to launch, simulated data is necessary- Section 2.3). We also quantify the sensitivity of our algorithm to the expected measurement error implicit in SWOT data. Finally, we compare our algorithm against published methods for estimating carbon dioxide evasion ($FCO_2$) from rivers: we use previously published dissolved $CO_2$ data to compare the bulk carbon efflux (via $CO_2$ evasion) from the 22 rivers as calculated using both our new algorithm and published gauged techniques.

##2 Methods
We name the RS of $k_{600}$ algorithm BIKER, or the ‘Bayesian Inversion of the $k_{600}$ Rate’ algorithm. To build BIKER, two steps are taken: first, we develop a $k_{600}$ upscaling model that relies on no a priori knowledge of the river (section 2.1) and then we couple that upscaling model with remote sensing techniques to build BIKER (section 2.2). Following these methods, the validation setup is described (section 2.3), as is the comparison of estimated bulk carbon effluxes from a suite of average flow velocity models (section 2.4). A flowchart detailing the complete workflow is included (Figure 1).

![Figure 1. Flowchart of methodology used in this study. ‘Extended Upscaling Model’ is detailed in section 2.1. ‘BIKER’ is detailed in section 2.2 and ‘Flux Estimation’ is detailed in section 2.4. See section 2.3 for the validation setup.](outputs/flowchart.jpg)

###2.1 Developing an RS-able $k_{600}$ upscaling model
[@raymondScalingGasTransfer2012a] first used river hydraulics and the theoretical relationship between *eD* and gas exchange to upscale $k_{600}$ from average flow velocity and channel slope, extending earlier work predicting the reaeration coefficient in a similar manner [e.g. @oconnorMechanismReaerationNatural1958; @tsivoglouTracerMeasurementReaeration1976]. Later, [@ulsethDistinctAirWater2019a] built upon [@raymondScalingGasTransfer2012a]’s work by expanding their dataset to include measurements from steeper rivers and finding that two distinct upscaling regimes exist in low and high energy rivers. More specifically, [@ulsethDistinctAirWater2019a] scale $k_{600}$ using *eD*, where the resulting upscaling model parameters are significantly different whether *eD* is high or low. Their upscaling function is reprinted as equation 2.

@@

While equation 2 would be ideal for the purposes outlined for this manuscript and provides best-to-date predictive accuracy for upscaling $k_{600}$, it requires an a priori knowledge of *eD* to assign its differential model parameters. We do not have this a priori knowledge in an ungauged, RS-based setting and only have access to SWOT-observed river widths and heights (and therefore water surface slopes). Thus, we need to extend the [@ulsethDistinctAirWater2019a] model to assign the two model parameters using only RS-able hydraulics. To do this, we use the basic premise that the $k_{600}\sim eD$ scaling relationship is different for different river sizes. We use river width to define river size and train different upscaling functions for different groups of data with similar river sizes [Figure 2, using the data from @ulsethDistinctAirWater2019a]. In doing so, we find that narrow rivers (< 10m wide) need to be further broken into low energy (slope < 0.05) and high energy (slope > 0.05) regimes in order to meet the assumptions necessary to fit a linear regression model (specifically, normally distributed residuals of the predictions).

@@

This model therefore assigns $a_i$ and $b_i$ a priori using just SWOT measurements of water surface width and slope to define *i*. Another implication of this extension of the [@ulsethDistinctAirWater2019a] model is that we now use *a* and *b* parameters that implicitly reflect the distinct geomorphology of different rivers.

@@

Finally, we estimated $σ_{upscale}^2$ using Monte Carlo (MC) methods to approximate total uncertainty from equation 4. Specifically, We explore three uncertainty scenarios to assess which source is most influential. Uncertainty in equation 4 can stem from three sources: 1) error baked into the upscaling model ($a_i$ and $b_i$ parameter uncertainty), 2) error in Manning’s approximation of the average flow velocity, and 3) measurement error. Our interest is using BIKER for satellite remote sensing and so we consider measurement errors from SWOT-derived water surface slopes. To do this, we run 15,000 unique MC simulations on 5,000 sets of field measurements of river channel hydraulics under three different uncertainty scenarios [@brinkerhoffReconcilingAtaStationAtManyStations2019]. Each MC simulation is itself 10,000 runs, sampling from the normal distributions for $a_i$ and $b_i$ (uncertainty scenario 1), $a_i$, $b_i$, and ln-transformed velocity (uncertainty scenario 2), and $a_i$, $b_i$, ln-transformed velocity, and water surface slope (uncertainty scenario 3). Estimate uncertainty for velocity followed @hagemannBAMBayesianAMHGManning2017a, who estimated ln-transformed Manning’s equation error to be 0.25. The specifics of the slope-error model used are elaborated in Section 2.3 [@durandHowWillRadar2020a]. This results in three sets of 5,000 discrete distributions for $k_{600}$ from which the median uncertainty term is extracted and used for $σ_{upscale}^2$.

With the sampling model (equation 5), prior distributions, and the $σ_{upscale}^2$ parameter described, a joint posterior distribution conditional on the SWOT data is described. To approximate this distribution, BIKER uses a Markov Chain Monte Carlo (MCMC) algorithm. Because it is written in the Stan probabilistic programming language, BIKER’s MCMC of choice is a Hamiltonian Monte Carlo which reduces computation time relative to other sampling algorithms (https://mc-stan.org/docs/2_26/reference-manual/hamiltonian-monte-carlo.html).

@@

Because SWOT has yet to launch, it is standard practice to benchmark SWOT-related algorithms on ‘SWOT-like data’. There are three current types of SWOT-like data: 1) AirSWOT, which is an airborne Ka-band inSAR currently limited to five rivers globally, 2) simulated rivers that mimic the type of data SWOT will provide, and 3) the SWOT simulator, which introduces measurement errors to these simulated rivers. Because we are principally interested in algorithm performance, we limit our validation setup to simulated rivers in order to benchmark across as many rivers as possible. These simulated rivers are simply reach-averaged hydraulic model outputs where the water surface heights and widths are labelled as ‘RS observations’ and are used as the sole inputs to the BIKER algorithm. Simulated rivers mimic perfect measurement conditions (i.e. no measurement errors). Here, we use the 22 rivers archived by @frassonCompilationHydraulicModels2019a and used by @rodriguezObservingRiversVarying2020 to develop reach-averaged Saint Vernant equations for the SWOT mission. We omitted the Arial Khan river in Bangladesh due to known problems with that hydraulic model and we further sampled the models for only the observations every 11 days to mimic the average SWOT overpass frequency (this also approximately aligns the SWOT observations with the bi-weekly $CO_2$ data used later in this study- Section 2.4). All told, this yielded 503 sets of SWOT observations to validate BIKER with.

We also introduced measurement error into these simulated SWOT observations to assess BIKER's performance degradation due to errors inherent in SWOT measurements. Error in SWOT measurements will come from both the error tolerances intrinsic in the satellite data product as well as radar layover error. Here, we assume errors in river width are negligible [@durandHowWillRadar2020a] and solely focus on measurement errors in river height and slope. Layover error is the phenomenon when radar returns from different places arrive at the sensor at the same time, SOMETHING HERE. This generally occurs in areas of high topographic variation [@reesPhysicalPrinciplesRemote2013] where the land surface 'obscures' the water surface relative to SWOT's incidence angle. While the SWOT simulator mentioned above is ideal for this analysis, it is cumbersome to run on many rivers and so here we coarsely model measurement error using the results from a recent global anlaysis of SWOT radar layover errors [@durandHowWillRadar2020a]. They developed a first-principles model for layover error from SWOT observations and, using global hydrography, quantified SWOT measurement error from both error tolerances and radar layover for 220,924 river reaches. The 68th percentiles of these reach-scale errors were 1.7cm/km and 10.4cm for water surface slope and height, respectively. To apply this model to our 22 rivers, we generate new river heights *H* and water surface slopes $S_h$ by sampling every observation from normal distributions following equations 7 and 8. Because BIKER ingests *dA*, rather than *H*, reach-scale height errors were converted to *dA* errors, again following [@durandHowWillRadar2020a] and our rectangular channel assumptions: $\sigma_{dA} \approx W\sigma_H\sqrt{2}$.

@@

Regardless of the validation data used, we do not have observed $k_{600}$ data for these rivers, and to our knowledge no field dataset of $k_{600}$ exists in the type of temporal and spatial frequency that SWOT (and therefore the BIKER algorithm) provides. Further, we are principally interested in our ability to reproduce upscaled $k_{600}$ from RS observations and are less concerend with the actual accuracy of the upscaling model itself [which can be validated using existing datasets like previously done by @raymondScalingGasTransfer2012a and @ulsethDistinctAirWater2019a]. Therefore, we take the best performing upscaling model from sections 2.1 and 3.1 and use that to calculate the ‘observed’ $k_{600}$ that BIKER is validated against. This is done using equation 9, where $V_{obs} [L/T]$ is observed river discharge divided by observed channel area. With this setup, we can directly explore BIKER's ability to infer observed *V* and $k_{600}$ from river width and height alone. It also means that, for a fair validation scheme, $\sigma_{upscale}^2$ must be set to reflect only error from Manning’s equation (and not $a_i$ and $b_i$). **PROBABLY NEED TO ACCOUNT FOR SLOPE UNCERTAINITY TOO**. Thus, $\sigma_{upscale}^2$ is set to 0.25 for this validation. However in practice, it should reflect the total uncertainty calculated in section 2.2 from all three sources.

@@

##3 Results
First, we present the results from the $k_{600}$ upscaling experiment (section 3.1). We then take the best performing upscaling function, implement it within BIKER, and validate it on the 22 SWOT rivers (section 3.2). Finally, we use $k_{600}$ as modeled by both BIKER and gauge-based HG models to compare $FCO_2$ and bulk carbon efflux estimates from the 22 rivers across gauged and ungauged methods (section 3.3).

###3.1 Validating the $k_{600}$ upscaling model
Figure 3 plots the validations for the five upscaling functions tested. Both the original [@ulsethDistinctAirWater2019a] model and our RS-able extension perform the best on this independent set of validation data, with near identical performance. The RS-able [@ulsethDistinctAirWater2019a] model achieves an $r^2$ of `r results_3_1[results_3_1$Model == 'RS-able',]$r2` and an RMSE of `r results_3_1[results_3_1$Model == 'RS-able',]$rmse` m/day, which is just marginally better than the original [@ulsethDistinctAirWater2019a] model (`r results_3_1[results_3_1$Model == 'Ulseth',]$r2` and `r results_3_1[results_3_1$Model == 'Ulseth',]$rmse` m/day, respectively). The two power-law based [@raymondScalingGasTransfer2012a] models (equations 4 and 3 in that study) achieve slightly worse performance, with $r^2$ in the 0.70s and their RMSE scores slightly higher (`r results_3_1[results_3_1$Model == 'Raymond_2',]$rmse` and `r results_3_1[results_3_1$Model == 'Raymond_3',]$rmse` m/day, respectively). [@raymondScalingGasTransfer2012a] equation 5, which is linear in structure, worked well on their dataset but fails to capture the non-linear relationship evident in this expanded and more geomorphically diverse dataset. In fact, it predicts negative $k_{600}$ values and error metrics cannot be calculated because the prediction residuals are not normally distributed. **SHOULD I INCLUDE THIS MODEL??** Because we successfully reproduced the [@ulsethDistinctAirWater2019a] model, the rest of this manuscript deals exclusively with this RS-able variant of the model.
